Moisture absorption and drying alter nonadditive litter flammability in a mixed conifer forest

### 2020-05-22 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Add comment with "tweetable abstract ?!"
Edits to MS in response to reviews on R2

\nwords{6330 (main text only), 8000 (abstract, text, captions) }
\ntables{4}
\nfig{5}
\nref{57}

@@

@@

      \item \emph{Synthesis:} This study is the first to examine ecologically
    relevant drying rates and their interaction with non additive misture
    effects on flammability. We find that taxon specific patterns of moisture
    absorption and drying change flammability rankings under moist compared
    with dry fuels because traits that increase flame spread rate increase
    water absorption. Litter mixtures generally burn with fire behavior more
    similar to that of the most flammable constituent species and this effect
    increases as fuels dry.

@@

In mixed species communities, species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such nonadditive effects are often synergistic in dry litter fuels
where the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
nonadditivity has also been recorded for litter mixtures
\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible
that similar nonadditive effects may influence litter moisture content and
flammability of moist fuels. There is some evidence for nonadditive
flammability in moist fuels with a single study finding increasing nonadditivity
with moisture \citep{Blauw+Wensink+etal-2015}, but this previous work did not
include estimation of the dry-down process and had an unclear relationship to
fire behavior in the field. In mixed conifer forest in California, where the
fire regime is characterized by surface fires fueled by plant litter, overstory
tree composition drives variation in local fire behavior
\citep{Schwilk+Caprio-2011} as a result of differential litter flammability
across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
flammability feedbacks may help explain historical fire frequency patterns
\citep{Schwilk+Caprio-2011}, but current estimates of litter flammability in
mixed conifer forests are based largely on burning dry litter.

This study is the first to examine ecologically relevant drying rates and their
interaction with non additive moisture effects on flammability. We asked if
differences in litter moisture absorption and drying rates change the
relative flammability of litter across species compared with flammability
measurements based on dry fuels. We investigated these litter moisture dynamics
across eight tree species in a temperate fire-prone forest in California. We
determined the maximum water absorption and rate of moisture loss and examined
how these traits vary in relation to specific leaf area and litter packing
density. Experimental burn trials of leaf litter at various moisture levels
determined how time since wetting influenced two axes of flammability
characterized by flame spread rate and total fuel consumed. To determine the
potential magnitude of mixture effects, we tested if the nonadditive effects on
reported in previous studies occur under higher moisture levels, and whether
fuel moisture effects strengthen or moderate any nonadditivity.

@@

Across these species there were significant differences in the maximum water
absorption and drying rate (all $P < 0.0001$, Fig.~\ref{fig:drydown},
Table~\ref{tab:drydown} and SI Figs.~1 -- 3). Calculating estimated marginal
means following the linear mixed-effects model discriminated the species into
three different maximum water absorption groups: 1) \emph{Q. kelloggii}; 2)
\emph{P. lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica},
\emph{S. giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P.
  ponderosa} (SI Fig.~2). Estimated mixed linear model slopes discriminated
species into three groups from lowest to most rapid drying rate: 1) \emph{C.
  decurrens}, 2) \emph{Abies spp.} and \emph{S. giganteum}, and 3) \emph{Q.
  kelloggii} and \emph{Pinus spp.} (SI Fig.~3).

\begin{figure}
  \centering
  \includegraphics[width=8cm]{figs_tables/fig1_dry_effects_by_taxon.pdf}
  \caption{}
\end{figure}
%%\clearpage
\begin{figure}
  \captionsetup{labelformat=adja-page}
  \ContinuedFloat
\caption{Effects of litter drying on fuel moisture and flammability. A) Dry
  down curves for eight species. Lines show best-fit exponential decay curves
  fit to moisture content on a dry mass basis. B) Flame spread rate by time
  since wetting. Lines indicate best fit linear model fits by taxonomic group.
  Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI
  Table 6. C) Percent fuel consumed by time since wetting. Curves shown
  indicate best fit beta regression lines by taxonomic group. Model ANOVA table
  in Table \ref{tab:consume_anova} and coefficients in SI Table 7. For all
  plots, horizontal jitter (0.5 to 2 hr) was added to avoid overplotting. A
  lack of data at earlier time points for a taxon indicates no ignitions
  occurred at those time points. }
 \label{fig:drydown}
\end{figure}
%% \clearpage

@@

Across these eight species, specific leaf area had a positive association with
maximum water absorption (Table~\ref{tab:mc_di_anova}, SI Fig~4, linear model
$P = 0.003$) and litter bulk density had a negative effect on drying rate
(Table~\ref{tab:mc_di_anova}, SI Fig~5, linear model $P = 0.029$). Past work
has demonstrated that green leaf length predicts litter packing and,
consequently, flame spread rate across these species. Litter particle size
distributions, however show that there is substantial variation in particle
sizes within a species (SI Fig.~6).

@@

Flame spread rate declined with fuel moisture (Fig.~\ref{fig:flam_moist}) and
maximum rates and the steepness of decline differed by taxonomic group
(Table~\ref{tab:spreadrate_anova}). Similarly, fuel moisture negatively
influenced fuel consumption (Fig.~\ref{fig:flam_moist},
Table~\ref{tab:consume_anova}). Although the overall ranking of groups was
similar for fuel consumption and for flame spread rate, there were different
pattern across taxa in these two flammability parameters. For spread rate, the
decline in flammability with moisture was steepest for \emph{Quercus} and
shallowest for \emph{Pinus} and Cupressaceae (Fig.~\ref{fig:flam_moist}). For
fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in
flammability with fuel moisture than did the other groups
(Fig.~\ref{fig:flam_moist}).

%% spread rate by moisture
\begin{figure}
  \centering
\includegraphics[width=8cm]{figs_tables/fig2_moisture_flam.pdf}
\caption{Fuel moisture effects on flammability. A) Flame spread rate by
  moisture content (model fit $F = 61.15$, df = 9, 148, $P < 0.0001$). Model
  ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI Table
  4. B) Percent fuel consumed by moisture content. Lines indicate best fit beta
  regression lines. Model deviance table in Table \ref{tab:consume_anova} and
  coefficients are in SI Table 5.}
  \label{fig:flam_moist}
\end{figure}

@@

As a consequence of the relationships between fuel moisture and flammability,
time since wetting had a significant effect on spread rate ($P < 0.0001$) and
the slope of this effect varied by taxonomic group (Fig. \ref{fig:drydown},
Table~\ref{tab:spreadrate_anova}). The pines were able to ignite and burn much
sooner after wetting than did the other taxa, none of which had measurable
spread rates until burning was attempted 72 hours after wetting. Trials that
failed to even ignite smoldering combustion are omitted completely from the
model and figure. At 72 hours, the \emph{Abies}, \emph{Calocedrus}, and
\emph{Sequoiadendron} had drier litter than did the \emph{Quercus}, yet the
spread rates were slow. \emph{Quercus} had the steepest response to time since
wetting (Fig.~\ref{fig:drydown}, SI Table~6)).

Fuel consumption was highly variable across trials and intermediate values were
less common than were very low or moderately high consumption
(Fig.~\ref{fig:drydown}). All parameters significantly contributed to the
generalized linear mixed beta regression model as indicated by the nested model
comparisons in Table~\ref{tab:consume_anova}. As was the pattern for flame
spread rate, pines had rapid flammability increases with drying. However,
in contrast to the other flammability analyses, here the pines did not behave
consistently: \emph{P. lambertiana} had lower fuel consumption than did \emph{P.
  ponderosa} and \emph{P. jeffreyi}. We noticed this during trials and in model
fitting we found that placing \emph{P. lambertiana} in its own group allowed
model convergence that was otherwise impossible. This was at odds with our
intent to use only \emph{a priori} taxonomic groups, but we could not obtain
model convergence otherwise. This result should therefore be seen as
preliminary because the test does not reflect an \emph{a priori} hypothesis,
but probably reflects the fact that \emph{P. lambertiana} has shorter and finer
needles than the other two pines.

@@

\begin{figure}
  \centering
\includegraphics[width=16cm]{figs_tables/fig3_mixture_obs_vs_pred_mc.pdf}
\caption{Observed vs predicted fuel moisture content for litter mixtures.
  Predictions are based on the mean moisture content at each time since wetting
  for the three constituent species in each mixture. Codes for species in
  mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi =
  \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_mc}
\end{figure}

@@

\begin{figure}
  \centering
\includegraphics[width=16cm]{figs_tables/fig4_mixture_obs_vs_pred_spread.pdf}
\caption{Observed vs. predicted flame spread rate for litter mixtures. Filled
  circles show observed spread rate values. Open diamonds indicate predicted
  spread rates based on the average spread rate at each time since wetting for
  the three constituent species in each mixture. Codes for species in mixtures:
  Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi =
  \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_spread}
\end{figure}

\begin{figure}
  \centering
\includegraphics[width=16cm]{figs_tables/fig5_mixture_obs_vs_pred_consume.pdf}
\caption{Observed vs predicted fuel consumption for litter mixtures. Filled
  circles show observed fuel consumption percentages. Open diamonds indicate
  predicted fuel consumption based on the average fuel consumption at each
  particular time since wetting for the three constituent species in each
  mixture. Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca =
  \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu =
  \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_consume}
\end{figure}

Flammability measures also demonstrated nonadditivity in mixtures. Flame spread
rates were generally higher than predicted (Fig.
\ref{fig:mixture_obs_pred_spread}, Wald test of mixed linear model centered
intercept for observed-predicted values greater than zero, $P = 0.017$) and the
positive nonadditivity increased with time since wetting (slope greater than
zero, $P = 0.005$). Fuel consumption, however, showed no overall bias towards
positive or negative nonadditivity, but nonadditivity increased with time since
wetting (Fig. \ref{fig:mixture_obs_pred_consume}, slope not equal to zero,
$P < 0.0001$). Fuel consumption tended to show negative nonadditivity at higher
fuel moistures and positive nonadditivity following drying
(Fig.~\ref{fig:mixture_obs_pred_consume}).

@@

The taxa in this study displayed different patterns of both moisture absorption
and desorption, associated with different litter traits. Those species that had
a greater absorptive capacity were also among the most flammable in terms of
spread rate when dry. Traits such as large leaves and high SLA may affect both
absorptive capacity and flammability with a consequence that flammability
rankings across species can change with time since wetting. One shortcoming of
past work in this system was that flammability trials tested only dry fuels
\citep{Magalhaes+Schwilk-2012}. We therefore expected that rankings of species
flammability might change under higher moisture levels. For example, one
presumption of ours prior to this experiment was that the high flammability
previously reported for \emph{Q. kelloggii} \citep{Magalhaes+Schwilk-2012}
would lessen significantly under higher fuel moisture. This prediction
was partly supported: \emph{Q. kelloggii} fails to ignite until after 72 hours
of drying in this study (Figs. \ref{fig:drydown} and \ref{fig:drydown}).
However, high moisture content in this oak was unable to suppress ignition of
mixtures to which it contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
\ref{fig:mixture_obs_pred_consume}), indicating the importance of non-additive
effects.

@@

Overall, we report a consistent pattern of increasingly positive nonadditivity
in flammability with litter drying across two largely orthogonal measures of
flammability \citep{Schwilk-2015, Pausas+Keeley+etal-2017,
  Prior+Murphy+etal-2018}. Some mixtures, however did exhibit negative
nonadditivity. For example, at 72 hours following saturation, fuel consumption
across mixtures showed neutral, positive, or negative nonadditivity (Fig.
\ref{fig:mixture_obs_pred_consume}). Previous studies have demonstrated the
potential for the direction of nonadditivity to shift with fuel moisture
\citep{Blauw+Wensink+etal-2015} or with altering mixtures of leaf vs twig
litter \citep{Zhao+vanLogtestijn+etal-2019}. Our litter represented naturally
dropped litter and the ratio of leaves to twigs in such material can vary both
by species but also by wind and storm events the previous season. With only
four distinct litter combinations, it is impossible to attribute mechanism to
the trend. Litter packing alone is a complex trait: needles and twigs can
combine such that the addition of a species to a mixture can create a denser
litter bed or less dense litter bed than is expected
\citep{Zhao+vanLogtestijn+etal-2019}.

We draw two main conclusions from this work. First, maximum water absorption is
highest in those species with the most flammability litter under dry conditions
(higher SLA, larger leaf area, less dense litter). As a consequence, under
higher fuel moisture, the flammability ranking of these taxa can change
compared to the flammability of dry fuels (for example, crossing lines in Fig.
\ref{fig:flam_moist} B and C). Second, litter flammability is prone to
nonadditive effects which become more positive with drying. The relationships
we described here can form a working hypothesis and baseline expectation to
adjust fire behavior predictions in real communities: in mixtures with fuel
moisture content below 50\%, observed flame spread rates approached double the
predicted value based on monoculture trials (Fig.
\ref{fig:mixture_obs_pred_spread}). In completely dry litter from this same
system, observed flame spread rates in mixtures approached three times the
expectation based on the average of constituent species
\citep{Magalhaes+Schwilk-2012}. These numbers provide a starting point for
parameterizing fire models with mixed fuels. In this temperate mixed conifer
forest, these effects are likely to include positive feedbacks whereby the
shade intolerant species such as the long-needled pines and the large-leaved
oak, are favored by the increased fire spread to which their fuels contribute
\citep{Schwilk+Caprio-2011}. Similar results are reported for southeastern US
forests where species composition shifts have led to decreases in fire activity
\citep{Nowacki+Abrams-2008}. Vegetation-fire feedbacks are important at a wide
range of scales from stand-level species composition shifts to the earth system
\citep{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}. The patterns we
demonstrate here show the importance of species traits and interactions at
local and stand scales where management and fire operations meet species'
biology.

@@

\section*{Data and code accessibility}
All data and code for this project is at
\url{https://github.com/schwilklab/trait-flam}. DOI to be provided. For review,
use ``joe'' branch which contains this manuscript and code needed to create the
figures and tables.

>>>>

\nwords{6330 (main text only), 8000 (abstract, text, captions) }
\ntables{4}
\nfig{3}
\nref{57}

@@

% Tweetable:

% Moisture dynamics changes flammability rankings. Litter mixtures nonadditivel and this effect increases as fuels dry.

% 50 word summary

% This study examined ecologically relevant drying rates and their
% interaction with on litter flammability. We demonstrated that fuel
% moisture changes flammability and interactions in mixed
% communities: litter mixtures generally burn with fire behavior
% more similar to that of the most flammable constituent species
% and this effect increases as fuels dry.

@@

      \item \emph{Synthesis:} This study is the first to examine ecologically
    relevant drying rates and their interaction with nonadditive mixture
    effects on flammability. We find that taxon specific patterns of moisture
    absorption and drying change flammability rankings under moist compared
    with dry fuels because traits that increase flame spread rate increase
    water absorption. Litter mixtures generally burn with fire behavior more
    similar to that of the most flammable constituent species and this effect
    increases as fuels dry.

@@

In mixed species communities, species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such nonadditive effects are often synergistic in dry litter fuels
where the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
nonadditivity has also been recorded for litter mixtures
\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible
that similar nonadditive effects may influence litter moisture content and
flammability of moist fuels. There is some evidence for nonadditive
flammability in moist fuels with a single study finding increasing
nonadditivity with moisture \citep{Blauw+Wensink+etal-2015}, but this previous
work did not include estimation of mixture effect on the dry-down process and
had an unclear relationship to fire behavior in the field. In mixed conifer
forest in California, where the fire regime is characterized by surface fires
fueled by plant litter, overstory tree composition drives variation in local
fire behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
flammability across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
flammability feedbacks may help explain historical fire frequency patterns
\citep{Schwilk+Caprio-2011}, but current estimates of litter flammability in
mixed conifer forests are based largely on burning dry litter.

This study is the first to examine ecologically relevant drying rates and their
interaction with nonadditive moisture effects on flammability. We asked if
differences in litter moisture absorption and drying rates change the
relative flammability of litter across species compared with flammability
measurements based on dry fuels. We investigated these litter moisture dynamics
across eight tree species in a temperate fire-prone forest in California. We
determined the maximum water absorption and rate of moisture loss and examined
how these traits vary in relation to specific leaf area and litter packing
density. Experimental burn trials of leaf litter at various moisture levels
determined how time since wetting influenced two axes of flammability
characterized by flame spread rate and total fuel consumed. To determine the
potential magnitude of mixture effects, we tested if the nonadditive effects on
reported in previous studies occur under higher moisture levels, and whether
fuel moisture effects strengthen or moderate any nonadditivity.

@@

Across these species there were significant differences in the maximum water
absorption and drying rate (all $P < 0.0001$, Fig.~\ref{fig:drydown}A,
Table~\ref{tab:drydown} and SI Figs.~1 -- 3). Calculating estimated marginal
means following the linear mixed-effects model discriminated the species into
three different maximum water absorption groups: 1) \emph{Q. kelloggii}; 2)
\emph{P. lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica},
\emph{S. giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P.
  ponderosa} (SI Fig.~2). Estimated mixed linear model slopes discriminated
species into three groups from lowest to most rapid drying rate: 1) \emph{C.
  decurrens}, 2) \emph{Abies spp.} and \emph{S. giganteum}, and 3) \emph{Q.
  kelloggii} and \emph{Pinus spp.} (SI Fig.~3).

\begin{figure}
  \centering
  \includegraphics[width=16cm]{figs_tables/fig1_dry_effects_by_taxon.pdf}
  \caption{}
\end{figure}
%%\clearpage
\begin{figure}
  \captionsetup{labelformat=adja-page}
  \ContinuedFloat
\caption{Effects of litter drying on fuel moisture and flammability. A) Dry
  down curves for eight species. Lines show best-fit exponential decay curves
  fit to moisture content on a dry mass basis. B) Flame spread rate by time
  since wetting. Lines indicate best fit linear model fits by taxonomic group.
  Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI
  Table 6. C) Percent fuel consumed by time since wetting. Curves shown
  indicate best fit beta regression lines by taxonomic group. Model ANOVA table
  in Table \ref{tab:consume_anova} and coefficients in SI Table 7.
  D) Flame spread rate by moisture content (model fit $F = 61.15$, df = 9, 148,
  $P < 0.0001$). Model ANOVA table in Table \ref{tab:spreadrate_anova} and
  coefficients in SI Table 4. E) Percent fuel consumed by moisture content.
  Lines indicate best fit beta regression lines. Model deviance table in Table
  \ref{tab:consume_anova} and coefficients are in SI Table 5. For plots in left
  column, horizontal jitter (0.5 to 2 hr) was added to avoid overplotting. A
  lack of data at earlier time points for a taxon indicates no ignitions
  occurred at those time points.}
 \label{fig:drydown}
\end{figure}
%% \clearpage

@@

Across these eight species, specific leaf area had a positive association with
maximum water absorption (Table~\ref{tab:mc_di_anova}, SI Fig~4, linear model
$P = 0.003$) and litter bulk density had a negative effect on drying rate
(Table~\ref{tab:mc_di_anova}, SI Fig~5, linear model $P = 0.029$). Past work
has demonstrated that green leaf length predicts litter packing and,
consequently, flame spread rate across these species
\citep{Magalhaes+Schwilk-2012}. Litter particle size distributions, however
show that there is substantial variation in particle sizes within a species (SI
Fig.~6).

@@

Flame spread rate declined with fuel moisture (Fig.~\ref{fig:drydown}D) and
maximum rates and the steepness of decline differed by taxonomic group
(Table~\ref{tab:spreadrate_anova}). Similarly, fuel moisture negatively
influenced fuel consumption (Fig.~\ref{fig:drydown}E,
Table~\ref{tab:consume_anova}). Although the overall ranking of groups was
similar for fuel consumption and for flame spread rate, there were different
pattern across taxa in these two flammability parameters. For spread rate, the
decline in flammability with moisture was steepest for \emph{Quercus} and
shallowest for \emph{Pinus} and Cupressaceae (Fig.~\ref{fig:drydown}D). For
fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in
flammability with fuel moisture than did the other groups
(Fig.~\ref{fig:drydown}E).

@@

As a consequence of the relationships between fuel moisture and flammability,
time since wetting had a significant effect on spread rate ($P < 0.0001$) and
the slope of this effect varied by taxonomic group (Fig. \ref{fig:drydown},
Table~\ref{tab:spreadrate_anova}). The pines were able to ignite and burn much
sooner after wetting than did the other taxa, none of which had measurable
spread rates until burning was attempted 72 hours after wetting. Trials that
failed to even ignite smoldering combustion are omitted completely from the
model and figure. At 72 hours, the \emph{Abies}, \emph{Calocedrus}, and
\emph{Sequoiadendron} had drier litter than did the \emph{Quercus}, yet the
spread rates were slow. \emph{Quercus} had the steepest response to time since
wetting (Fig.~\ref{fig:drydown}B, SI Table~6)).

Fuel consumption was highly variable across trials and intermediate values were
less common than were very low or moderately high consumption
(Fig.~\ref{fig:drydown}C and E). All parameters significantly contributed to the
generalized linear mixed beta regression model as indicated by the nested model
comparisons in Table~\ref{tab:consume_anova}. As was the pattern for flame
spread rate, pines had rapid flammability increases with drying. However,
in contrast to the other flammability analyses, here the pines did not behave
consistently: \emph{P. lambertiana} had lower fuel consumption than did \emph{P.
  ponderosa} and \emph{P. jeffreyi}. We noticed this during trials and in model
fitting we found that placing \emph{P. lambertiana} in its own group allowed
model convergence that was otherwise impossible. This was at odds with our
intent to use only \emph{a priori} taxonomic groups, but we could not obtain
model convergence otherwise. This result should therefore be seen as
preliminary because the test does not reflect an \emph{a priori} hypothesis,
but probably reflects the fact that \emph{P. lambertiana} has shorter and finer
needles than the other two pines.

@@

\begin{figure}
  \centering
\includegraphics[width=16cm]{figs_tables/fig2_mixture_obs_vs_pred_mc.pdf}
\caption{Observed vs predicted fuel moisture content for litter mixtures.
  Predictions are based on the mean moisture content at each time since wetting
  for the three constituent species in each mixture. Codes for species in
  mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi =
  \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_mc}
\end{figure}

@@

\begin{figure}
  \centering
\includegraphics[width=16cm]{figs_tables/fig3_mixture_obs_vs_pred_flam.pdf}
\caption{Observed vs. predicted flammability for litter mixtures. Filled
  circles show observed values. Open diamonds indicate predicted values based
  on the average spread or fuel consumption at each time since wetting for the
  three constituent species in each mixture. Codes for species in mixtures: Ab
  = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus
    jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_flam}
\end{figure}

Flammability measures also demonstrated nonadditivity in mixtures. Flame spread
rates were generally higher than predicted (Fig.
\ref{fig:mixture_obs_pred_flam}A, Wald test of mixed linear model centered
intercept for observed-predicted values greater than zero, $P = 0.017$) and the
positive nonadditivity increased with time since wetting (slope greater than
zero, $P = 0.005$). Fuel consumption, however, showed no overall bias towards
positive or negative nonadditivity, but nonadditivity increased with time since
wetting (Fig. \ref{fig:mixture_obs_pred_flam}B, slope not equal to zero,
$P < 0.0001$). Fuel consumption tended to show negative nonadditivity at higher
fuel moistures and positive nonadditivity following drying
(Fig.~\ref{fig:mixture_obs_pred_flam}B).

@@

The taxa in this study displayed different patterns of both moisture absorption
and desorption, associated with different litter traits. Those species that had
a greater absorptive capacity were also among the most flammable in terms of
spread rate when dry. Traits such as large leaves and high SLA may affect both
absorptive capacity and flammability with a consequence that flammability
rankings across species can change with time since wetting. One shortcoming of
past work in this system was that flammability trials tested only dry fuels
\citep{Magalhaes+Schwilk-2012}. We therefore expected that rankings of species
flammability might change under higher moisture levels. For example, one
presumption of ours prior to this experiment was that the high flammability
previously reported for \emph{Q. kelloggii} \citep{Magalhaes+Schwilk-2012}
would lessen significantly under higher fuel moisture. This prediction was
partly supported: \emph{Q. kelloggii} fails to ignite until after 72 hours of
drying in this study (Figs. \ref{fig:drydown} and \ref{fig:drydown}). However,
high moisture content in this oak was unable to suppress ignition of mixtures
to which it contributed (Fig. \ref{fig:mixture_obs_pred_flam}), indicating the
importance of nonadditive effects.

@@

Overall, we report a consistent pattern of increasingly positive nonadditivity
in flammability with litter drying across two largely orthogonal measures of
flammability \citep{Schwilk-2015, Pausas+Keeley+etal-2017,
  Prior+Murphy+etal-2018}. Some mixtures, however did exhibit negative
nonadditivity. For example, at 72 hours following saturation, fuel consumption
across mixtures showed neutral, positive, or negative nonadditivity (Fig.
\ref{fig:mixture_obs_pred_flam}B). Previous studies have demonstrated the
potential for the direction of nonadditivity to shift with fuel moisture
\citep{Blauw+Wensink+etal-2015} or with altering mixtures of leaf vs twig
litter \citep{Zhao+vanLogtestijn+etal-2019}. Our litter represented naturally
dropped litter and the ratio of leaves to twigs in such material can vary both
by species but also by wind and storm events the previous season. With only
four distinct litter combinations, it is impossible to attribute mechanism to
the trend. Litter packing alone is a complex trait: needles and twigs can
combine such that the addition of a species to a mixture can create a denser
litter bed or less dense litter bed than is expected
\citep{Zhao+vanLogtestijn+etal-2019}.

We draw two main conclusions from this work. First, maximum water absorption is
highest in those species with the most flammability litter under dry conditions
(higher SLA, larger leaf area, less dense litter). As a consequence, under
higher fuel moisture, the flammability ranking of these taxa can change
compared to the flammability of dry fuels (for example, crossing lines in Fig.
\ref{fig:drydown} D and E). Second, litter flammability is prone to
nonadditive effects which become more positive with drying. The relationships
we described here can form a working hypothesis and baseline expectation to
adjust fire behavior predictions in real communities: in mixtures with fuel
moisture content below 50\%, observed flame spread rates approached double the
predicted value based on monoculture trials (Fig.
\ref{fig:mixture_obs_pred_flam}A). In completely dry litter from this same
system, observed flame spread rates in mixtures approached three times the
expectation based on the average of constituent species
\citep{Magalhaes+Schwilk-2012}. These numbers provide a starting point for
parameterizing fire models with mixed fuels. In this temperate mixed conifer
forest, these effects are likely to include positive feedbacks whereby the
shade intolerant species such as the long-needled pines and the large-leaved
oak, are favored by the increased fire spread to which their fuels contribute
\citep{Schwilk+Caprio-2011}. Similar results are reported for southeastern US
forests where species composition shifts have led to decreases in fire activity
\citep{Nowacki+Abrams-2008}. Vegetation-fire feedbacks are important at a wide
range of scales from stand-level species composition shifts to the earth system
\citep{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}. The patterns we
demonstrate here show the importance of species traits and interactions at
local and stand scales where management and fire operations meet species'
biology.

@@

\section*{Data and code accessibility}
All data and code for this project is at
\url{https://github.com/schwilklab/trait-flam}. Checkout git tag ``JOE.R3''
which contains this manuscript and code needed to create the figures and
tables.
